VOI-aware Monte-Carlo Sampling in Trees

### 2011-12-10 paper/vago/paper.tex
typo fixed

The simple regret of a sampling policy for the Multi-armed Bandit
Problem is the expected difference between the best expected reward
$\mu_*$ and the expected reward $\mu_j$ of the arm with the best sample mean
$\overline X_j=\max_i\overline X_i$:
\begin{equation}
\label{eqn:simple-regret}
\IE[R]=\sum_{j=1}^K\Delta_j\Pr(\overline X_j=\max_i\overline X_i)
\end{equation}
where $\Delta_j=\mu_*-\mu_j$.
Strategies that minimize the simple regret are called pure exploration
strategies \cite{Bubeck.pure}. Principles of rational metareasoning
\cite{Russell.right} suggest that at each step the arm with the great
value of information (VOI) must be pulled, and the sampling must be
stopped and a decision must be made when no arm has positive VOI.

@@

Pure exploration in multi-armed bandit problems is explored in
\cite{Bubeck.pure}. On the one hand, the paper proves certain upper
and lower bounds for UCB1 and uniform sampling, showing that an upper
bound on the simple regret is exponential in the number of samples for
uniform sampling, while only polynomial for UCB1. On the other hand,
empirical performance of UCB1 appears to be better than of uniform
sampling. \cite{Mnih.bernstop} investigate stopping criteria for
sampling based on the empirical Bernstein inequality; however, the
stopping criteria are based on error probabilities rather than on
value of information measures, and do not directly address the
objective of regret minimization.

@@

Let $X_1, \ldots, X_n$ be i.i.d. random variables with values from $[0,1]$,
$X=\frac 1 n \sum_{i=1}^n X_i$. Then 
\begin{description}
\item[Hoeffding inequality \rm{\cite{Hoeffding.ineq}}:] 
\begin{equation}
\Pr(X-\IE[X] \ge a) \le \exp ( -2na^2)
\label{eqn:conc-hoeffding}
\end{equation}
\item[Empirical Bernstein inequality \rm{(derived in Appendix~\ref{app:deriv-conc-empbernstein})}:]
\begin{eqnarray}
\Pr(X-\IE[X] \ge a) &\le& 2\exp \left( - \frac {na^2} {\frac {14} {3}
                          \frac {n} {n-1}a+2\overline\sigma_n^2}\right)\nonumber\\
                    &\le& 2\exp \left( - \frac {na^2} {10a+2\overline\sigma_n^2}\right)
\label{eqn:conc-empbernstein}
\end{eqnarray}
where sample variance $\overline\sigma_n^2$ is
\begin{equation}
\overline\sigma_n^2=\frac 1 {n(n-1)} \sum_{1\le i < j\le n}(X_i-X_j)^2
\label{eqn:sample-variance}
\end{equation}
\end{description}
Bounds (\ref{eqn:conc-hoeffding}, \ref{eqn:conc-empbernstein}) are symmetrical
around the mean. Bound~(\ref{eqn:conc-empbernstein}) is tighter than
(\ref{eqn:conc-hoeffding}) for small $a$ and $\overline\sigma_n^2$.

@@

The search time is finite, and in a simple case the \textit{search
  budget} specified as the maximum number of samples.
An estimate based on the perfect intrinsic VOI does not take in
consideration the remaining number of samples. Given two arms
with the same intrinsic perfect VOI, the VOI
estimate of the arm pulled fewer times so far should be higher.

@@

An upper bound on the intrinsic value of perfect information is obtained by substituting
(\ref{eqn:probound-perf-hoeffding}) into (\ref{eqn:thm-vopi}):
\begin{equation}
  \Lambda_i^p \le \hat\Lambda_i^p=\left\{
  \begin{array}{l l}
    \overline X_\beta\exp\left(- 2(\overline X_i - \overline X_\beta)^2 n_i \right) & \mbox{{\rm if} $i=\alpha$} \\
    (1-\overline  X_\alpha)\exp\left(- 2(\overline X_\alpha - \overline X_i)^2 n_i \right)(1-\overline  X_\alpha)  &  \mbox{\rm otherwise}
  \end{array} \right.
\label{eqn:bound-perf-hoeffding}
\end{equation}
An upper bound on the blinkered VOI estimate with ordering independent of $N$ is obtained
by substituting (\ref{eqn:probound-blnk-hoeffding}) into (\ref{eqn:thm-be}).
\begin{equation}
  \Lambda_i^b \le \hat\Lambda_i^b=\left\{
  \begin{array}{l l}
    N \frac {\overline X_\beta} {n_i}\exp\left(- 1.37(\overline X_i - \overline X_\beta)^2 n_i\right)
      & \mbox{{\rm if} $i=\alpha$} \\
    N \frac {(1-\overline  X_\alpha)} {n_i} \exp\left(- 1.37(\overline X_\alpha - \overline X_i)^2 n_i\right)
      &  \mbox{\rm otherwise}
  \end{array} \right.
\label{eqn:bound-blnk-hoeffding}
\end{equation}
The bound is generally tighter when $N$ is small relative to $n_i$,
and the estimate $\hat\Lambda_i^b$ can be viewed as a relaxation of the myopic VOI
estimate to longer sequences of actions, such
that the VOI estimate is positive for some of the actions.

@@

A natural approach would be to apply an algorithm that minimizes the
simple regret in Multi-armed bandits at the first step, and the
cumulative regret (solves the exploration-exploitation trade-off)
during the rest of the rollout.  However, this approach assumes that
the information obtained from rollouts in the current state is
discarded after an action is selected. In practice, most successful
Monte-Carlo tree search algorithms re-use rollouts originating from an
earlier search state and passing through the current search state;
thus, the value of information of a rollout is determined not just by
the influence on the choice of the action at the current state, but
also by the influence on the choice at future search states, provided
the rollout passes through the search states which will be visited.

@@

The sampling policies are first compared on random multi-armed bandit 
problem instances, where the Monte-Carlo Tree Search algorithm
is reduced to the sample selection policy at the first step: UCT is 
actually UCB1. In the problem of pure exploration in multi-armed bandits
\cite{Bubeck.pure}, the smaller is the exploration factor, the higher
is the expected simple regret; indeed, UCT exhibits the lowest simple
regret in the experiments for the default exploration factor $Cp=\frac
{\sqrt 2} 2$ \cite{Kocsis.uct}.

Figure~\ref{fig:random-instances} shows experiment results for
randomly-generated multi-armed bandits with 32 Bernoulli arms, with
the mean rewards of the arms distributed uniformly in the range $[0,
  1]$, for a range of sample budgets $32..2048$, with multiplicative
step of $2$. The experiment for each number of samples was repeated
20000 times. UCT is worse than any of the VOI-aware sampling policies,
for example, for 384 samples the simple regret of UCT is $\approx 40$
times greater than of VCT, and $\approx 100$ times greater than of
ECT. On average, it takes $2.5$ times as many samples for UCT to reach
the regret of VCT. Different numbers of arms and distributions of
means give similar results.

@@

This work suggested Monte-Carlo sampling policies in which sample
selection and termination are based on upper bounds on the value of
information. A simplified model of accounting for future value of
information of a sample, based on early termination and sample
redistribution, was proposed for Monte-Carlo sampling in
trees. Empirical evaluation showed that the policies outperform
heuristic algorithms for simple regret minimization in multi-armed
bandits, as well as for tree search.

>>>>

The simple regret of a sampling policy for the Multi-armed Bandit
Problem is the expected difference between the best expected reward
$\max\limits_i\IE[X_i]$ and the expected reward $\IE[X_j]$ of the arm
with the best sample mean $\overline X_j=\max\limits_i\overline X_i$:
\begin{equation}
\label{eqn:simple-regret}
\IE[R]=\sum_{j=1}^K\Delta_j\Pr(\overline X_j=\max_i\overline X_i)
\end{equation}
where $\Delta_j=\max\limits_i\IE[X_i]-\IE[X_j]$.  Strategies that
minimize the simple regret are called pure exploration strategies
\cite{Bubeck.pure}. Principles of rational metareasoning
\cite{Russell.right} suggest that at each step the arm with the great
value of information (VOI) must be pulled, and the sampling must be
stopped and a decision must be made when no arm has positive VOI.

@@

Pure exploration in Multi-armed bandits is explored in
\cite{Bubeck.pure}. On the one hand, the paper proves certain upper
and lower bounds for UCB1 and uniform sampling, showing that an upper
bound on the simple regret is exponential in the number of samples for
uniform sampling, while only polynomial for UCB1. On the other hand,
empirical performance of UCB1 appears to be better than of uniform
sampling. \cite{Mnih.bernstop} investigate stopping criteria for
sampling based on the empirical Bernstein inequality; however, the
stopping criteria are based on error probabilities rather than on
value of information measures, and do not directly address the
objective of regret minimization.

@@

Let $X_1, \ldots, X_n$ be i.i.d. random variables with values from $[0,1]$,
$X=\frac 1 n \sum_{i=1}^n X_i$. Then 
\begin{description}
\item[Hoeffding inequality \rm{\cite{Hoeffding.ineq}}:] 
\begin{equation}
\Pr(X-\IE[X] \ge a) \le \exp ( -2a^2n)
\label{eqn:conc-hoeffding}
\end{equation}
\item[Empirical Bernstein inequality \rm{(derived in Appendix~\ref{app:deriv-conc-empbernstein})}:]
\begin{eqnarray}
\Pr(X-\IE[X] \ge a) &\le& 2\exp \left( - \frac {a^2n} {\frac {14} {3}
                          \frac {n} {n-1}a+2\overline\sigma_n^2}\right)\nonumber\\
                    &\le& 2\exp \left( - \frac {a^2n} {10a+2\overline\sigma_n^2}\right)
\label{eqn:conc-empbernstein}
\end{eqnarray}
where sample variance $\overline\sigma_n^2$ is
\begin{equation}
\overline\sigma_n^2=\frac 1 {n(n-1)} \sum_{1\le i < j\le n}(X_i-X_j)^2
\label{eqn:sample-variance}
\end{equation}
\end{description}
Bounds (\ref{eqn:conc-hoeffding}, \ref{eqn:conc-empbernstein}) are symmetrical
around the mean. Bound~(\ref{eqn:conc-empbernstein}) is tighter than
(\ref{eqn:conc-hoeffding}) for small $a$ and $\overline\sigma_n^2$.

@@

The search time is finite, and in a simple case the \textit{search
  budget} is specified as the maximum number of samples.
An estimate based on the perfect intrinsic VOI does not take in
consideration the remaining number of samples. Given two arms
with the same intrinsic perfect VOI, the VOI
estimate of the arm pulled fewer times so far should be higher.

@@

An upper bound on the intrinsic value of perfect information is obtained by substituting
(\ref{eqn:probound-perf-hoeffding}) into (\ref{eqn:thm-vopi}):
\begin{equation}
  \Lambda_i^p \le \hat\Lambda_i^p=\left\{
  \begin{array}{l l}
    \overline X_\beta\exp\left(- 2(\overline X_i - \overline X_\beta)^2 n_i \right) & \mbox{{\rm if} $i=\alpha$} \\
    (1-\overline  X_\alpha)\exp\left(- 2(\overline X_\alpha - \overline X_i)^2 n_i \right)  &  \mbox{\rm otherwise}
  \end{array} \right.
\label{eqn:bound-perf-hoeffding}
\end{equation}
An upper bound on the blinkered VOI estimate with ordering independent of $N$ is obtained
by substituting (\ref{eqn:probound-blnk-hoeffding}) into (\ref{eqn:thm-be}).
\begin{equation}
  \Lambda_i^b \le \hat\Lambda_i^b=\left\{
  \begin{array}{l l}
    N \frac {\overline X_\beta} {n_i}\exp\left(- 1.37(\overline X_i - \overline X_\beta)^2 n_i\right)
      & \mbox{{\rm if} $i=\alpha$} \\
    N \frac {(1-\overline  X_\alpha)} {n_i} \exp\left(- 1.37(\overline X_\alpha - \overline X_i)^2 n_i\right)
      &  \mbox{\rm otherwise}
  \end{array} \right.
\label{eqn:bound-blnk-hoeffding}
\end{equation}
The bound is generally tighter when $N$ is small relative to $n_i$,
and the estimate $\hat\Lambda_i^b$ can be viewed as a relaxation of the myopic VOI
estimate to longer sequences of actions, such
that the VOI estimate is positive for some of the actions.

@@

A natural approach would be to apply an algorithm that minimizes the
simple regret in Multi-armed bandits at the first step, and the
cumulative regret during the rest of the rollout.  However, this
approach assumes that the information obtained from rollouts in the
current state is discarded after an action is selected. In practice,
most successful Monte-Carlo tree search algorithms re-use rollouts
originating from an earlier search state and passing through the
current search state; thus, the value of information of a rollout is
determined not just by the influence on the choice of the action at
the current state, but also by the influence on the choice at future
search states, provided the rollout passes through the search states
which will be visited.

@@

The sampling policies are first compared on random Multi-armed bandit 
problem instances, where the Monte-Carlo Tree Search algorithm
is reduced to the sample selection policy at the first step: UCT is 
actually UCB1. In the problem of pure exploration in Multi-armed bandits
\cite{Bubeck.pure}, the smaller is the exploration factor, the higher
is the expected simple regret; indeed, UCT exhibits the lowest simple
regret in the experiments for the default exploration factor $Cp=\frac
{\sqrt 2} 2$ \cite{Kocsis.uct}.

Figure~\ref{fig:random-instances} shows experiment results for
randomly-generated Multi-armed bandits with 32 Bernoulli arms, with
the mean rewards of the arms distributed uniformly in the range $[0,
  1]$, for a range of sample budgets $32..2048$, with multiplicative
step of $2$. The experiment for each number of samples was repeated
20000 times. UCT is worse than any of the VOI-aware sampling policies,
for example, for 384 samples the simple regret of UCT is $\approx 40$
times greater than of VCT, and $\approx 100$ times greater than of
ECT. On average, it takes $2.5$ times as many samples for UCT to reach
the regret of VCT. Different numbers of arms and distributions of
means give similar results.

@@

This work suggested Monte-Carlo sampling policies in which sample
selection and termination are based on upper bounds on the value of
information. A simplified model of accounting for future value of
information of a sample, based on early termination and sample
redistribution, was proposed for Monte-Carlo sampling in
trees. Empirical evaluation showed that these policies outperform
heuristic algorithms for simple regret minimization in Multi-armed
bandits, as well as for tree search.
